---
title: "mrgsolve"
author: "Shen Cheng"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. `mrgsolve` in general
- an `R` package for simulation from hierarchical, ordinary differential equation (ODE) based models typically employed in drug development.
    - free, open-source software
    - utilizes an ODE-solver from `ODEPACK` (freely-available in the public domain)
- Language
    - Models written in `C++` inside model specification format
    - Simulation workflow in `R`
    - Integrate certain `NONMEM` functionality so easier for `NONMEM` users to get involved
        - `F`, `ALAG`, `SS`, `AMT`,`CMT`, etc

# 2. Installation
- [The trickest part](https://github.com/metrumresearchgroup/mrgsolve)
    - `install.packages("mrgsolve")`
    - `remotes::install_github("metrumresearchgroup/mrgsolve@dev")`
    - `C++` complier
    - `FORTRAN` complier
- [installation guide](https://github.com/metrumresearchgroup/mrgsolve/wiki/mrgsolve-Installation) 
- Prerequists:
    - `R`: https://cran.r-project.org/
        - Currently `R` >= 3.1.2 is required
    - Seriously consider running in `Rstudio` if possible. This is not required but is recommended.
    - Compilers: `C++` and `gfortran` 

# 3. Simulation with `mrgsolve`

Two major components:  

- Model specification file written in `C++`
- Simulation specification file written in `R`
- We are able to combine two files into one as a single `R` script or `R markdown` file

## 3.1 Model specification file

- The model file for `mrgsolve` is written in `C++` syntax and saved as a `.cpp` file.
- Alternatively, the model can be written in `C++` syntax and saved as a `R` object

```{r example model file}

one_comp_mod <- '

[set] delta = 0.01, end = 50 //set options for simulations

[PROB]
// $PROB
 - author: Shen Cheng
 - date: `r Sys.Date()`
 - comments: this is a demonstration
 - one compartment model + first order absportion
 - random effects added on `CL` and `VC` as exponential distributions
 - proportional error model
 - model covariate: SEX ~ CL

$PARAM
TVCL = 1, TVVC = 20, TVKA = 1, SEXCL=0.5, SEX=0

$CMT  @annotated 
// similar to $MODEL in NONMEM
// if @annotated is not used, simply declare compartment names with a space

DEPOT  : Depot compartment (mass)
CENT   : Central compartment (mass)

$MAIN
// similar to $PK block in NONMEM
  
double CL   = TVCL*exp(ETA(1))*pow(SEXCL, SEX); //power function in C++ witten as pow(,)
double VC   = TVVC*exp(ETA(2))                ; // "double" means defining the data structure as double
double KA   = TVKA                            ; // ";" is required at the end of each defined variable

  
$GLOBAL 
// declare variables you want to either use further in computation
// OR derive variables you want calculated from computations

// bool cure = FALSE; // declaring variables needs ";" at the end
// double CL, VC, KA; // declaring variables so no "double" is need in $MAIN
#define CP (CENT/VC)  // # define does not require ";"

//[PREMABLE]
// similar to $GLOBAL to some extent
// However, unlike $MAIN $TABLE and $ODE, $PREMABLE only call once during the simulation

$OMEGA @block
// @block generates a block matrix with the formulation (1,1), (1,2), (2,2)....
// @correlation will specify values to be corrleations instead as variances

0.1 0.02 0.3

$SIGMA 
// sigma variances go here

0.01

$ODE
//similar to $DES in NONMEM
//differential equations go here
  
dxdt_DEPOT = -KA*DEPOT;
dxdt_CENT  = KA*DEPOT - CL*CP;

$TABLE
//another block could be used for defining secondary parameters

double DV = CP * (1 + EPS(1));

$CAPTURE
//This is a block to identify variables that should be captured in the simulated output

CP, DV

'
```

## 3.2 Simulation specification file

### 3.2.1 Read model in R mrgsolve

```{r call library, message=FALSE}
library(mrgsolve) 
library(tidyverse) #simulation with mrgsolve is normally integrated in a tidyverse workflow (lots of use of " %>% ")
library(ggpubr) #package for combining plots
```

* We can read in our model from a saved `.cpp` file using `mread`  
    -`mread("file name", "path")`

```{r mread}
mod <- mread("one_comp_mod", "/Users/shencheng/Documents/GitHub/SummerSession2021/Session2_Simulations")
```

* or we can read in our model from a `R` object using `mcode`
    -`mcode("model name", code)`

```{r mcode, eval=FALSE}
mod <- mcode("one_comp_mod", one_comp_mod)
```

### 3.2.2 Model check

```{r model check}
mod        #model check
param(mod) #check model parameters
init(mod)  #check initial conditions of model compartments
```

### 3.2.3 Simulation workflow

* A commonly used simulation workflow could be something like the codes shown below
    
```{r simulation work flow}
#an illustration of the common workflow of simulation using mrgsolve
mod %>%                           #start with model object
  #req(CP, DV) %>%                 #request specific model output 
  ev(amt=100, cmt=1, time=0) %>%  #specify the events
  mrgsim(end=100, delta=0.1) %>%  #starts simulations, end and delta can be specified in the model file or within mrgsim 
  plot()                          #a quick simulation check
  #plot(CP~time)                  #only plot CP vs time
```

* `nid` within `mrgsim` allows us to quickly do a population simulation

```{r simulate population profiles}
mod %>%                           
  req(CP, DV) %>%                  
  ev(amt=100, cmt=1, time=0) %>%          
  mrgsim(end=100, delta=0.1, nid=10) %>% #nid: specify number of subjects we want to simulate 
  plot() 
```

* Simulation with a dataset using `data_set()` could be very useful
     - a dataset normally includes both interventions and populations

```{r simulation with a dataset}
data <- data.frame(ID = seq(1, 10), #simulate 10 subjects
                   amt = 100, #amt specify doses
                   cmt = 1, #specify dosing compartment
                   time = 0, #specify dosing time
                   evid = 1, #this is a doing event 
                   SEX = sample(c(0,1), replace=TRUE, size=10)) #model covariate SEX, 0 means female, 1 means male
                                                                #sample(), take 10 samples from c(0,1) with replacement 

mod %>%                           
  req(CP, DV) %>%                  
  data_set(data) %>%   #simulate from a dataset, a dataset typically include event and population       
  mrgsim(end=100, delta=0.1) %>%  
  plot() 
```

* `plot()` after `mrgsim()` allows a quick check of the model output
* Plotting with `ggplot()` allows more flexiblity and higher quality

```{r plot with ggplot}
sim <- 
  mod %>%                           
  req(CP, DV) %>%                  
  data_set(data) %>%       
  mrgsim(end=100, delta=0.1) %>% 
  as.data.frame()  #turn model simulation into a data frame

sim %>% 
  ggplot(aes(x=time, y=CP, group=ID, color=as.character(ID)))+
  geom_line()+
  xlab("Time")+
  ylab("Concentration")+
  theme_bw()  #plot with ggplot 
```

* Sometimes we may want to simulate without random effects. 
* `zero_re()` sets all random effects to 0 when model compile in

```{r simulation without random effect}
mod <- 
  mread("one_comp_mod", "/Users/shencheng/Documents/GitHub/SummerSession2021/Session2_Simulations") %>% 
  zero_re() # sets random effects to 0 when model compile in
```

```{r check omega matrix}
omat(mod)  #check model omega matrix
smat(mod)  #check model sigma matrix
```


* Let's simulate 10 subjects and check their difference

```{r quick check of model simulation}
mod %>%                           
  req(CP, DV) %>%                  
  ev(amt=100, cmt=1, time=0) %>%          
  mrgsim(end=100, delta=0.1, nid=10) %>% 
  plot() 
```

* Check sex effect on CL

```{r check covariate effects}
data_pop <- 
  data.frame(ID = seq(1, 2), #simulate 2 subjects
             amt = 100, #amt specify doses
             cmt = 1, #specify dosing compartment
             time = 0, #specify dosing time
             evid = 1, #this is a doing event 
             SEX = c(0,1)) #model covariate SEX, 0 means female, 1 means male

sim_pop <- 
  mod %>%                           
  req(CP, DV) %>%                  
  data_set(data_pop) %>%       
  mrgsim(end=100, delta=0.1) %>% 
  as.data.frame()  #turn model simulation into a data frame

sim_pop %>% 
  mutate(SEX = if_else(ID == 1, 0, 1)) %>% 
  mutate(SEX = as.character(SEX)) %>% 
  ggplot(aes(x=time, y=CP, group=ID, color=SEX))+
  geom_line()+
  xlab("Time")+
  ylab("Concentration")+
  theme_bw()
```

### 3.2.4 Interventions

* `ev()` or `data_set()` are commonly used for specify interventions. Here're the things that can be recognized as interventions in either `ev()` or `data_set()`:

    * `time` event time
    * `cmt` Event compartment
    * `amt` Dose amount
    * `ii` Inter-dose interval
    * `addl` Additional doses to administer
    * `rate` Infusion rate
        + rate = 0 is bolus
        + rate > 0 is zero order infusion (`evid` 1)
    * `ss` Set to 1 to advance to steady state
    * `evid` Event id. 
        + `evid` 0 is for "observations"
        + `evid` 1 is for "dose"
        + `evid` 2 is for "other"
        + `evid` 3 is for compartment reset
        + `evid` 4 is reset and dose
    * `ID` Subject ID (use multiple ids - ID = 1:10)

*`ev()` can easily combine of interventions

```{r combine interventions}
e1 = ev(amt=100, time=0, ii=24, addl=1)
e2 = ev(amt=50, time=48, ii=24, addl=2)
e3 = c(e1, e2)  #combine e1 and e2

e4 = ev(amt=100, ii=24, addl=1)
e5 = ev(amt=50, ii=24, addl=2)
e6 = seq(e4, wait=36, e5)  #combine e4 and e5 in sequence
                           #`wait=` sets intervals between interventions 

p1 <- mod %>% ev(e1) %>% mrgsim(delta=0.01, end=200) %>% plot(CP ~ time)
p2 <- mod %>% ev(e2) %>% mrgsim(delta=0.01, end=200) %>% plot(CP ~ time)
p3 <- mod %>% ev(e3) %>% mrgsim(delta=0.01, end=200) %>% plot(CP ~ time)
p4 <- mod %>% ev(e4) %>% mrgsim(delta=0.01, end=200) %>% plot(CP ~ time)
p5 <- mod %>% ev(e5) %>% mrgsim(delta=0.01, end=200) %>% plot(CP ~ time)
p6 <- mod %>% ev(e6) %>% mrgsim(delta=0.01, end=200) %>% plot(CP ~ time)

ggarrange(p1, p2, p3, p4, p5, p6, ncol=3, nrow=2)
```

* `expand_ev()` facilate the simulations of different dosing regimens

```{r expand.ev()}
data_ev1 <- expand.ev(amt=c(50, 100, 150), time=0)
data_ev2 <- expand.ev(amt=100, ii=c(12, 24, 36), addl=2)
data_ev3 <- expand.ev(amt=c(50, 100, 150), ii=c(12, 24, 36), addl=2) #`expand_ev` on multiple factors sets all combinations on test conditions  

p7 <- mod %>% data_set(data_ev1) %>% mrgsim(end=150, delta=0.1) %>% plot(CP~ time)
p8 <- mod %>% data_set(data_ev2) %>% mrgsim(end=150, delta=0.1) %>% plot(CP~ time)
p9 <- mod %>% data_set(data_ev3) %>% mrgsim(end=150, delta=0.1) %>% plot(CP~ time)

ggarrange(p7, p8, p9, ncol=3)
```


### 3.2.5 Sensitivity analysis with `mrgsolve`

* Sensitivity analysis is commonly used for assessing the impact of the parameters on model predictions. 
* `mrgsolve` enables multiple ways to conduct sensitivity analysis
    - `expand_idata`
    - `knobs()`

```{r expand.idata}
idata <- expand.idata(TVCL = c(0.5, 1, 1.5))

mod %>% 
  idata_set(idata) %>% 
  ev(amt=100, cmt=1, time=0) %>% 
  mrgsim(end=50, delta=0.5) %>% 
  plot()
```



```{r knobs()}
mod %>% 
  ev(amt=100, cmt=1, time=0) %>% 
  knobs(TVCL = c(0.5, 1, 1.5)) %>%  #no `mrgsim` is needed if `knobs` is used
  plot()
```


### 3.2.6 population simulation with `mrgsolve`

* Let's first read back in our model with random effects

```{r read in model with random effects again}
mod <- mread("one_comp_mod", "/Users/shencheng/Documents/GitHub/SummerSession2021/Session2_Simulations")
```

```{r confirm random effects are non-zero}
omat(mod)  #check model omega matrix
smat(mod)  #check model sigma matrix
```

```{r conduct a population simulations}
sim1 <- 
  mod %>% 
  ev(amt=100, cmt=1, time=0) %>% 
  mrgsim(end=100, delta=0.1, nid=100, obsonly=TRUE) %>% #`obsonly` will carryout only observation records (evid=0)
  as.data.frame()

sim1 %>% 
  ggplot(aes(x=time, y=CP, group=ID))+
  geom_line()+
  theme_bw()
```

* With functionalities in `tidyverse`, we can calculate summary statistics and then plot. 

```{r}
sim1 %>% 
  group_by(time) %>% 
  mutate(median = median(CP), 
         upper = quantile(CP, 0.975), 
         lower = quantile(CP, 0.025)) %>% 
  ggplot()+
  geom_line(aes(x=time, y=median), color="red", size=1.2)+
  geom_ribbon(aes(x=time, ymin=lower, ymax=upper), fill="gray20", alpha=0.25)+
  theme_bw()
```

### 3.2.7 update model parameters

* Model parameters can be updated on the fly

- Use `param()` for updating parameters in [PARAM] 

```{r}
param(mod)
mod2 <- mod %>% param(TVVC = 30)
param(mod2)
```

- Use `omat` to update parameters in [OMEGA]

```{r}
omat(mod)
mod3 <- mod %>% omat(bmat(0.1, 0.002, 0.3)) #bmat used for updating a block matrix
omat(mod3)
```

- Use `smat` to update parameters in [SIGMA]

```{r}
smat(mod)
mod4 <- mod %>% smat(dmat(0.02)) #dmat used for updating a diagnal matrix
smat(mod4)
```

- Evaluate the impact of the covariance between `ETA(1)` and `ETA(2)`

```{r}
mod5 <- mod %>% omat(dmat(0.1, 0.3))  #zero-out model covariance
mod6 <- mod %>% omat(bmat(0.1, 0.15, 0.3)) 

omat(mod5)
omat(mod6)
```

```{r, warning = FALSE}
pop_sim <- function(model){
  set.seed(0720)  #set a seed to facilitate the reproducibility of the simulation
  df <- 
    model %>% 
    ev(amt=100, cmt=1, time=0) %>% 
    mrgsim(end=100, delta=0.1, nid=100, obsonly=TRUE) %>% 
    as.data.frame()
  
  df <- 
    df %>% 
    group_by(time) %>% 
    mutate(median = median(CP), 
           upper = quantile(CP, 0.975), 
           lower = quantile(CP, 0.025)) %>% 
    ungroup()
  
  p <- 
    df %>% 
    ggplot()+
    geom_line(aes(x=time, y=median), color="red", size=1.2)+
    geom_ribbon(aes(x=time, ymin=lower, ymax=upper), fill="gray20", alpha=0.25)+
    scale_y_log10(limits = c(0.000000001, 15))+
    theme_bw()
}

p1 <- pop_sim(mod5) + ggtitle("cov = 0")
p2 <- pop_sim(mod6) + ggtitle("cov = 0.15")

ggarrange(p1, p2, ncol=2)
```

# 4. Useful Resources

* https://github.com/mrgsolve/learn
* GitHub site: https://github.com/metrumresearchgroup/mrgsolve
* mrgsolve website: https://mrgsolve.github.io
* User Guide: https://mrgsolve.github.io/user_guide


