---
title: "Normalized Prediction Distribution Errors"
author: "Mutaz M. Jaber"
date: " `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    number_sections: true
    md_extensions: +tex_math_single_backslash
    mathjax: https://cdnjs.cloudflare.com/ajax/libs/mathjax/3.2.0/es5/latest.js?tex-mml-chtml.js
  pdf_document:
    number_sections: true
    md_extensions: -tex_math_single_backslash
---


Normalized prediction distribution errors (npde) is an evaluation metric used
in nonlinear mixed-effect approach (NLME). This metric is a nonparametric
version of population weighted residual calculation. 


## General notation

Let $i$ denote the $i^{th}$ individual and $j$ is the $j^{th}$ measurment point
in an individual. $n_{tot}$ is the total observations in dataset.

$$
n_{tot} = \sum_{1}^i n_{i}
$$

Let $Y_i$ be the observation vector for subject i, and $f(.)$ is the PK model, then for the observation $y_{ij}$:

$$
y_{ij} = f(t_{ij}, \theta_i) + \epsilon_{ij} 
$$

Where $\theta_i$ is vector of individual PK parameters, that we assumed it is log-normally distributed

$$
\theta_i = h(\mu) e^{\eta_i}
$$ 

Where $\mu$ is the population value, and $\eta_i$ is the vector of random effects, and usually we assume that $\eta_i$ are normally distributed $\eta_i \sim \mathcal{N}(0, \omega^2)$ 

> Reminder: 

```{r, fig.align='center'}
set.seed(1234)
x <- rnorm(100,0, 2) 
hist(x, freq=F, xlab=expression(eta), main=expression(eta[i] %~~% N(0, 2)))
curve(dnorm(x, 0, 2),ylab='PDF',  add = TRUE)
```

## Predictive Distribution 

Taking into consideration that population model $f$ and population parameters
$\{\mu, diag(\Omega), \sigma\}$ are given and they define the null model
($H_0$).  We can define the predictive distribution ($pd_i$) for the $i^{th}$ subject as:

$$
p_i = \int p(y|\theta_i)p(\theta_i)d\theta_i
$$

Where $p(y|\theta_i)$ is normally distributed with mean $f(t_{ij}, \theta_i)$ and variance of $\Sigma_i(\theta_i) = diag(\sigma^2h(t_{ij}, \theta_i)$. Since we almost deal with nonlinear systems, there is no analytical expression for $p_i(y)$. Thus, we approximate the distribution using discrete distribution using stochastic simulation MCMC. 

$$
p_i(y) = \frac{1}{K} \sum_{k=1}^K p(y|\theta_k) = \frac{1}{K} \sum_{k=1}^K \phi(y; f(t_{ij}, \theta_k), \Sigma_i(\theta_k))
$$

Where $\phi$ is the multivariate normal PDF with mean of $\mu$ and variance of $\Omega$ evaluated at population model value. Prediction distribution can compare the observed data with the theoritical distribution of model prediction based on MCMC simulation. 

## Prediction Discrepancy 

Prediction discrepancy is a method uses stochastic MCMC simulation to generate prediction distribution for each observation. If the model and population parameters are correct, its expected that prediction discrepancy follows a uniform distribution: $pd_i \sim U(0,1)$. 


Let $F_{ij}$ be the prediction [distribution
function](https://en.wikipedia.org/wiki/Cumulative_distribution_function) of
the observation $y_{ij}$ for the $i^{th}$ under the estimated model $M$. We can
calculate prediction discrepancy $pd_{ij}$ as the value of $F_{ij}$ at
observation $y_{ij}$.    

We can simulate $K$ datasets and let $Y^{sim(k)}_i$ denotes the vector of simulated
observations for the $i^{th}$ subject in the $k^{th}$ dataset. $pd_i$ is computed as the percentile of $y_{ij}$ based on the theoritcal distribution of $y^{sim(k)}_{ij}$ 

$$
pd_{ij} = F_{ij}(y_{ij}) \approx \frac{1}{K} \sum \delta_{ijk}
$$

Where $\delta_{ijk}$ is an indicator function 

$$
\delta_{ijk} = \begin{cases} 1 & y_{ij}^{sim(k)} < y_{ij} \\ 0 & o.w. \end{cases} 
$$

However, as we mentioned the distribution of $pd$ is uniform but that the case of one observation per indiviudal without taken into consideration with-subject correlation of multiple observation. 

To overcome this issue, a decorrelation step has been proposed


$$
pde_{ij} = F^*_{ij}(y^*_{ij}) \approx \frac{1}{K} \sum \delta^*_{ijk}
$$
 
## Computation of NPDE

Applying the inverse normal distribution function to $pde_{ij}$ we can obtain normalized prediction distribution error ($npde_{ij}$):

$$
npde_{ij} = \phi^{-1}(pde_{ij})
$$

With that, if $H_0$ is true, them NPDE will follow a standard normal distribution $npde \sim \mathcal{N}(0, 1)$




## NONMEM example

Lets go back to PK hello world;IV-bolus one-compartment model with linear elimination. The implementation is simple in NONMEM, few parameters are added to `$TABLE` block for postprocessing

```
$TABLE ... NPDE NPD ESAMPLE=n
```

The full model: 

```
$PROB base 
; Drug X; one compartment
; estimat
$INPUT ID TIME AMT DROP DV MDV NUM
$DATA ../../../data/dat1.csv ignore=@   

$SUBR ADVAN1 TRANS2
$EST MET=1 PRINT=5 NOABORT INTER 

$PK
    TVCL = THETA(1) 
    TVV  = THETA(2) 

    CL  = TVCL * EXP(ETA(1))
    V   = TVV * EXP(ETA(2)) 

    S1 = V 

$ERROR

IPRED = F 
Y  = IPRED * (1 + ERR(1)) 

$THETA 
(0, 5) ; CL 
(0, 20) ; V

$OMEGA 
0.1
0.1

$SIGMA 0.01
$COVR 
$TAB ID NUM TIME AMT DV MDV IPRED PRED CWRES NPDE NPD ESAMPLE=10000 NOPRINT NOAPPEND ONEHEADER FILE=r2.tab 
```

Where `ESAMPLE` is the $K$ simulated datasets using MCMC. The default value is 300. 


We can plot the data for correlated NPDE and decorrelated one

```{r}
dat <- read.table('src/r1.tab', skip=1, header=TRUE)
dat <- dat[dat$CWRES!=0, ]
with(dat, plot(TIME, NPDE)) 
abline(h=0)
```


Now with population prediction 

```{r}
with(dat, plot(PRED, NPDE))
abline(h=0)
```

Same with correlated normalized prediction 

```{r}
with(dat, plot(TIME, NPD))
abline(h=0)
```

And again with population prediction

```{r}
with(dat, plot(PRED, NPD))
abline(h=0)
```


References:

- Computing normalised prediction distribution errors
to evaluate nonlinear mixed-effect models:
The npde add-on package for R [https://pubmed.ncbi.nlm.nih.gov/18215437/]
- NONMEM VIII User guide 
